### 2015-03-27 cug_paper/01_introduction.tex
Paper:  Lots of work on the CUG paper.

Discuss the general idea and the rationale behind it

>>>>

\todo[inline]{Discuss the general idea and the rationale behind it}

Titan, the Cray XK7 at Oak Ridge National Laboratory, has 18,688 compute nodes. Each node consists of a 16-core AMD CPU, an NVIDIA GPU and 32GB ram. In addition, there is another 6GB of ram on each GPU card. Not surprisingly, there are applications that do not use all the resources on a compute node.

Specifically, approximately one third of the applications that run on Titan do not use the GPU at all.\todo{Find a citation for this.}
There may be good reasons for this.  Utilizing the GPU's requires changes to the application and it's possible the developers haven't had the time to do that.  It's also possible that the type of computations the application performs simply don't lend themselves to GPU-based acceleration.  Regardless of why, it's clear that not all applications are benefiting from the available hardware.

\todo[inline]{Abrupt transition: we go from talking about hardware to talking about software...}

In most HPC applications, filesystem writes typically follow a very 'bursty' pattern. That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output when the application writes its checkpoint or output files. At ORNL, the filesystem is a shared resource and other users will also be putting a varying load on it. This means that the performance that a user sees at any given time will likely vary dramatically.

Having a sufficiently large write-back cache could improve applications' write performance in two ways.  First, if the cache is large enough to hold the entire write and the individual writes are far enough apart, then the application never has to wait on the filesystem.  It can write to the cache and then resume its computations while the cache drains to the filesystem.  Secondly, write performance should be more predictable. From the application's point of view, write speed would only limited by how fast data can be copied into the cache, and that is not effected by outside influences.

 The GPU memory and integer cores can be used to implement such a cache without taking resources away from the main application.

Since the GPU memory is accessed via the PCIe bus, it is generally not practical to use it as part of a process's main memory. Its bandwidth is too low and its latency is too high. However, when compared to filesystem performance - even a very fast parallel filesystem - accessing this memory is still very fast. Tests on Titan show that writing to GPU memory is at least eight times faster than writing to the filesystem, and considerably better than that if the filesystem is under heavy load. This suggests that using this memory as a large write-back cache for file I/O might help application runtimes significantly.

\begin{comment}

Specifically, though AMD markets the processor as having 16 cores, it actually has only 8 floating-point units. As such, some applications that are floating-point intensive run with only 8 processes per node. This actually leaves 8 cores available so long as they primarily execute integer instructions. Furthermore, some applications do not make use of the GPU and thus it, and the ram attached to it, also go unused.

Since the GPU memory is accessed via the PCIe bus, it is generally not practical to use it as part of a process's main memory. Its bandwidth is too low and its latency is too high. However, when compared to filesystem performance - even a very fast parallel filesystem - accessing this memory is still very fast. Tests on Titan show that writing to GPU memory is at least eight times faster than writing to the filesystem, and considerably better than that if the filesystem is under heavy load. This suggests that using this memory as a large write-back cache for file I/O might help application runtimes significantly.

In HPC applications, most filesystem writes follow a very ?bursty? pattern. That is, there will be long periods with no file output while the application is in a compute phase, followed by a short period of intense output when the application writes its checkpoint or output files. Furthermore, at ORNL, the filesystem is a shared resource and other users will also be putting a varying load on it. This means that the performance that a user sees at any given time will likely vary dramatically.

Having a sufficiently large write-back cache would alleviate both of these problems. Assuming the cache is large enough, then the 'bursty' nature of the output pattern can be smoothed out since there will be time for the traffic to drain from the cache before the next burst arrives. Equally importantly, write performance should be more predictable. From the application's point of view, write speed is only limited by how fast data can be copied into the cache, and that is not effected by outside influences. The GPU memory and integer cores can be used to implement such a cache without taking resources away from the main application.

This paper demonstrates a technique for providing such a write-back cache to applications. We use a synthetic benchmark to demonstrate performance without such a cache, with a cache that utilizes regular system ram and with a cache that utilizes ram on the GPU card. Performance benefits and any potential trade-offs and caveats are discussed.
\end{comment}

### 2015-03-27 cug_paper/02_design.tex
Paper:  Lots of work on the CUG paper.

Discuss how the software works.  Things like using CCI for small messages, bulk transfers happening by way of cudaMemcpy().  Mention Cray-specific stuff such as setting 'feature=gpudefault' in the qsub script.

>>>>

\todo[inline]{Discuss how the software works.  Things like using CCI for small messages, bulk transfers happening by way of cudaMemcpy().  Mention Cray-specific stuff such as setting 'feature=gpudefault' in the qsub script. }

\subsection{Overview}
\label{subsec:overview}
This idea was tested using a synthetic benchmark application written by the authors.  The benchmark consists of two executables: the main executable which simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.  The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.  Note that the daemon does \emph{not} copy the data into the GPU memory.  There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.  \footnote{Due to the limitations of aprun, starting this daemon is somewhat involved.  See section \ref{subsec:cray-specific}.}

With the daemon \todo{Should 'daemon' be plural?} started, the application begins its main execution.  In this case, since it's just a synthetic 
I/O benchmark, it performs no actual computation; it simply writes a specified amount of data and then sleeps for a specified amount of time.  For these tests, the sleep time was deliberately calculated to allow enough time for the data in GPU memory to drain.

When it's time to write, the application sends a short message to the daemon.  This message is a request to write data, and contains the amount of data and the starting offset into the output file.\footnote{For this benchmark, the daemon names output files based on the MPI rank of the requestor.  For production code, something more flexible will obviously be needed.}
\todo[inline]{We need a diagram showing the communication between the processes.}  The daemon will attempt to allocate memory on the GPU.  The daemon will then reply to the application with the size of memory that was actually allocated and a memory handle that the application can use to access the GPU memory.  Note that the size value may be less than what the application asked for.

Upon receiving the reply, the application opens the CUDA memory handle and copies the specified amount of data up to the GPU memory.  When the copy has completed, the application closes the memory handle and sends a message back to the daemon say that this particular write has completed.  If the application has more data to write, it can send another write request, or else it can continue with its calculations.

Note that the data has not made it out to the filesystem yet; it's just sitting in GPU memory.  Upon receiving the 'write done' message from the application, the daemon now knows it's free to

The daemon maintains one or more background threads that are responsible for copying data out of GPU memory into normal system memory and then writing that data to the filesystem.

The background write thread(s) operates asynchronously.

Each thread copies data out of GPU memory and into a buffer in system memory.  From there, it is written out to the filesystem.  The system memory buffer is deliberation kept fairly small - 16MB - in order to conserve resources.  Testing done on Titan shows that 16MB block sizes are large enough for cudaMemcpy() to give good performance.\cite{ac_guide}  Larger buffers would simply use up memory without improving performance noticeably.

\todo[inline]{Do we need to get into the details of cuda IPC functions? cudaIpcGetMemHandle, cudaIpcOpenMemHandle, cudaIpcCloseMemHandle}

From the application's point of view, the write has completed once the 'write done' message has been successfully sent.

Once the applications sends the 'write done' message

\todo[inline]{Need to mention CCI - short messages, low latency.  Citation reference is "atchley11:cci"}

The GPU attempts to perform some load-balancing by restricting the amount of memory allocated for any single request.  Currently this limit is 16MB.  If an application wants to write 512MB, it will therefore have to make 32 separate requests.  If other ranks on the node are trying to write at the same time, their requests will all be interleaved and hopefully rank will be starved.

It's debatable whether this sort of load balancing is necessary or even desirable.  Indeed, preliminary testing seemed to indicate that it might be faster to let the daemon allocate  as much GPU memory as the client requested (assuming there was enough free).  However, that same testing showed that one of the CUDA functions - cudaIpcOpenMemHandle() - became unreliable once GPU memory allocations exceeded 512MB.  Until the nature of that problem could be determined, it was decided to limit allocations to 16MB.

\subsection{Cray-Specific Details}
\label{subsec:cray-specific}

\todo[inline]{Need an introduction paragraph for this subsection.}
The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process or to multiple threads from multiple processes.  On Titan, if no special options are given to qsub, the CUDA runtime will be configured for a single thread.  Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.  This is done by passing the flag "-l feature=gpudefault" to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called 'gpudefault'.  The naming scheme comes from the CUDA documentation.  On a standard CUDA install, the GPU would in fact default to allowing access from multiple processes.  Titan is configured differently, however.  Thus the need to explicitly tell qsub to switch to the 'default' mode.}

\todo[inline] {talk about actually starting up the daemon - fork(), open(excl), exec() }

### 2015-03-27 cug_paper/03_analysis.tex
Paper:  Lots of work on the CUG paper.

\subsection{Experiment Setup}
Discuss what we ran, the parameters we swept and why we chose those parameters

\subsection{Results}
Discuss the results

>>>>

\subsection{Experiment Setup}
\label{subsec:exp_setup}
\todo[inline]{Discuss what we ran, the parameters we swept and why we chose those parameters}

Two experiments
First one: 8 ranks per node, 64 nodes.  The application ranks were placed on the even numbered cores and the daemon threads were pinned to the odd numbered cores. Write sizes varied from 1MB to 2GB and the daemon was run with 1, 2 \& 4 background threads.

Second experiment: 16 ranks per node, 64 nodes.  The application ranks consumed all cores on the node and the daemon threads were left free to be scheduled on any core.  The daemon was run with only a single write thread.  (Results from the first experiment showed no improvement when using multiple write threads, and on this test the cores were already oversubscribed.  Adding more write threads would only exacerbate that problem.)

Plots show the average node write speeds (i.e.: the sum of the write speeds reported by each rank running on the node) as seen from the perspective of the application.

\subsection{Results}
\label{subsec:results}
\todo[inline]{Discuss the results}

### 2015-03-27 cug_paper/04_futurework.tex
Paper:  Lots of work on the CUG paper.

Discuss work yet to be done:  add support for caching in system memory,  make 'production ready' (ie: easier for applications to use)

>>>>

\todo[inline]{Discuss work yet to be done:  add support for caching in system memory,  make 'production ready' (ie: easier for applications to use), measure/quantify any negative impacts on app performance (particularly for apps that run 16 ranks/node)}

\todo[inline]{ Summit will have node-local SSD's and that a variant of this code could be quite useful for applications that haven't been modified to take advantage of them more directly.  Not sure if we can mention Summit details yet.  Not sure if we should mention Summit at a Cray conference...  Maybe this paragraph should be in terms of generic burst-buffer support.}
